Week 1. Exploratory Data Analysis
=================================
Exploratory Graphs
-------------------
### One varialble
```{r}
pollution = read.csv('exploratoryGraphs/data/avgpm25.csv', colClasses = c('numeric','character','factor','numeric','numeric'))
head(pollution)
# quantile
summary(pollution)
# boxplot
boxplot(pollution$pm25, col = 'lightblue')
abline(h=12)
# histogram; add median and benchmark etc
hist(pollution$pm25, col = 'lightblue')
rug(pollution$pm25)
abline(v=12, lwd=2)
abline(v= median(pollution$pm25), col='purple', lwd=4)
hist(pollution$pm25, breaks = 100, col = 'lightblue')
rug(pollution$pm25)
# barplot with counts
barplot(table(pollution$region), col='wheat', main='number of counties in each region')
```

### Multivariable
```{r}
# flexible area plot
plot(table(pollution$region,pollution$pm25>12))
# boxplot, using ~ for conditioning
boxplot(pm25~region, data=pollution)
# hist, using par() for faceting
old.par = par()    # saving default par() first before changing
par(mfrow=c(2,1), mar=c(4,4,2,1))
hist(subset(pollution,region=='east')$pm25)
hist(subset(pollution,region=='west')$pm25)
par(old.par)    # reset to par default setting
# scatter plot, using with() & col
with(pollution, plot(latitude, pm25, col=region))
abline(h=12, lwd=2, lty=2)
# multi scatter plots
par(mfrow=c(1,2),mar=c(5,4,2,1))
with(subset(pollution,region=='west'), plot(latitude,pm25,main='West'))
with(subset(pollution,region=='east'), plot(latitude,pm25,main='East'))
par(old.par)
```

### Plotting System
Cant mix functions between different systems
* base: build up piece by piece
* lattice: build all in one go with single function call
* ggplot2: combine the merits of base and lattice, with layers as base and panels in lattice

### Base Plotting System
```{r}
# Base plot with annotation
# make May's points with a different color
with(airquality, plot(Wind, Ozone, main = 'Ozone and Wind in the NY City'))
with(subset(airquality, Month==5), points(Wind, Ozone, col='blue'))

# even fancier, with type='n' when initializing the plot
with(airquality, plot(Wind, Ozone, main='Ozone and Wind in NY'))
with(subset(airquality, Month==5), points(Wind, Ozone,col='blue'))
with(subset(airquality, Month!=5), points(Wind, Ozone,col='red'))
legend('topright', pch=1, col=c('blue','red'),legend=c('May','Other Months'))

# add lm model with abline()
with(airquality, plot(Wind, Ozone, main='Ozone and Wind in NY',pch=20))
model = lm(Ozone~Wind, airquality)
abline(model, lwd=2)

# multiple base plots with par() and mtext()
par(mfrow=c(1,3), mar=c(4,4,2,1), oma=c(0,0,2,0))
with(airquality, {
 plot(Wind, Ozone, main='Ozone and Wind')
 plot(Solar.R, Ozone, main='Ozone and Radiation')
 plot(Temp, Ozone, main='Ozone and Temperature')
 mtext('Ozone and Weather in NY City', outer=T)
})
```

### Graphics Devices in R
vector format:
* pdf: resize well, good for line type, not efficient if many points
* svg: xml-based, supports animation and interactivity

bitmap format:
* png: good for plotting many objects/points, bad for resize
* jped: lossy compression, good for many objects, bad for resize and line drawings

```{r}
# create mupltiple plots in mupltiple screens
dev.list()
dev.cur()
dev.set(3)
# copy plots from a plot device to a file device
## dev.copy
## dev.copy2pdf
par(old.par)
library(datasets)
with(faithful, plot(eruptions, waiting))
title(main='old faithful geyser data')
dev.copy(png, file='geyserplot.png')
dev.off()
```

